A Simple Type Theory for Category Theory

### 2022-04-15 paper/paper.tex
semantics work

In this section we establish our desired semantic soundness and
completeness theorems for models in virtual equipments with the
corresponding universal properties.
%
The first component of this is \emph{soundness}: every category,
object, set and transformation denote an object, vertical arrow,
horizontal arrow and 2-cell, respectively in any representable virtual
equipment, such that all equations in the theory are satisfied.
%
We combine this with a set theoretic semantics for types $A$ and
contexts $\Gamma$ to get a simple semantics of our logic over a
virtual equipment.
%
The second component of this is the \emph{completeness} theorem:
informally, the theory contains every category, object, set and
transformation that \emph{must} exist in any virtual equipment, and
further equations between syntactic forms hold if and only if they
hold in every model.
%
While sounding more ambitious, the completeness theorem is
typically simpler as it just amounts to showing that the syntax itself
forms a virtual equipment.

@@

%% \begin{definition}[Virtual Double Category]
%%   A virtual double category $V$ consists of
%%   \begin{enumerate}
%%     \item A set $V_0$ of objects.
%%     \item For each $\cat C, \cat D \in V_0$, a set $V_v(\cat C, \cat
%%       D)$ of vertical arrows, with composition and unit operations
%%       satisfying the category axioms.
%%     \item A set $V_h$ of horizontal arrows, with $d^+,d^- : V_h \to
%%       V_0$.
%%     \item For each composable string $\Phi$ of horizontal arrows
%%       starting at $\cat C$ and ending at $\cat D$, and horizontal
%%       arrow $R$, a set $V_2(\Phi; R)$ of 2-cells, with projection
%%       functions $d^{\pm} : V_2(\Phi;R) \to V_v(d^\pm \Phi, d^\pm R)$.

%%     \item For each $R \in V_h$, an identity arrow $\id_R \in
%%       V_2(d^-R,R,d^+ R; R)$ with $d^\pm \id_R = \id_{d^\pm R}$.

%%     \item For any $2$-cell $\alpha \in V_2(\Phi;R)$ and composable
%%       string $\phi$ of 2-cells into $\Phi$, a composite $\alpha \circ
%%       \phi$.

%%     \item Composition should be unital in that:
%%       \[ \id_R \circ [d^-\alpha, \alpha, d^+\alpha] = \alpha  \]
%%       and
%%       \[ \alpha \circ [\vec \id] = \alpha\]
%%     \item And associative in that
%%       \[ (\alpha \circ [\beta_1,\ldots]) \circ [\gamma_{11},\ldots,\gamma_{21},\ldots,] = \alpha \circ [(\beta_1 \circ \gamma_{11},\ldots), (\beta_2 \circ \gamma_{21},\ldots)]\]
%%   \end{enumerate}
%% \end{definition}

%% \begin{definition}[Restriction]
%%   Let $R \in V_h$ and $f \in V_v(\cat C, d^-R)$ and $g \in V_v(\cat D,
%%   d^+R)$, then a restriction of $R$ along $f,g$ is a horizontal arrow
%%   $R(f,g) \in V_h$ with $d^-R(f,g)= \cat C$ and $d^+R(f,g)= \cat D$
%%   with a (op?)-cartesian 2-cell $\alpha \in V_2(R(f,g),R)$ with
%%   $d^-\alpha = f$ and $d^+\alpha = g$.
%% \end{definition}

@@

\subsection{Category of Models}

Our goal is to establish that the syntax of SCT (with whichever type
constructors we choose) presents the initial indexed virtual equipment
(with the appropriate universal properties). To start we define
appropriate categories of equipments.

\begin{definition}{Indexed Virtual Equipment}
  A VE-hyperdoctrine is a functor $\cat C^{op} \to VE$, and a morphism
  of VE-hyperdoctrines is a natural transformation.
\end{definition}

  \begin{enumerate}
  \item The indexing cateogry is $\Set$.
  \item Given $\Gamma \in \Set$, the virtual equipment $\mathcal
    V^\Gamma$ is the ``power'' of $\mathcal V$ by $\Gamma$:
    \begin{enumerate}
    \item Objects are functions $\Gamma \to \mathcal V_0$.
    \item A vertical morphism from $O$ to $O'$ is a function $\Pi_{\gamma\in\Gamma} V_v(O\gamma,O'\gamma)$.
    \item A horizontal morphism is a function $\Gamma \to \mathcal V_h$. This satisfies
    \item A 2-cell from $R$ to $S$ is a family $\Pi_{\gamma\in\Gamma} \mathcal V_h(R\gamma,S\gamma)$.
    \end{enumerate}
    Compositions and restrictions are given pointwise.
  \item Restriction by a function $\gamma : \Delta \to \Gamma$ is
    given pointwise.
  \end{enumerate}
\end{definition}

\begin{definition}
  Given a well-formed signature $\Gamma$ and a virtual equipment
  $\mathcal V$, we mutually define the following notions:

  \begin{itemize}
  \item The set $\sem{\Gamma}_{\mathcal V}$
  \item For any $\Gamma \vdash A$ and $\gamma \in \sem{\Gamma}_{\mathcal V}$, a set $\sem{A}_{\mathcal V} \gamma$
  \item For any $\Gamma \vdash M : A$ and $\gamma \in \sem{\Gamma}_{\mathcal V}$, an element $\sem{M}_{\mathcal V} \in \sem{A}_{\mathcal V} \gamma$
  \item For any $\Gamma \vdash \cat C \isaCat$ and $\gamma \in \sem{\Gamma}_{\mathcal V}$, an object $\sem{\cat C}_{\mathcal V}\gamma$
  \item For any $\Gamma\pipe\alpha:\cat C \vdash a :\cat D$ and $\gamma \in \sem{\Gamma}_{\mathcal V}$, a vertical morphism $\sem{\alpha.a}_{\mathcal V}\gamma \in \mathcal V_v(\sem{\cat C}_{\mathcal V}\gamma, \sem{\cat D}_{\mathcal V}\gamma,)$
  \item For any $\Gamma\pipe\Xi \vdash R \isaSet$ and $\gamma \in
    \sem{\Gamma}_{\mathcal V}$, a horizontal arrow $\sem{R}_{\mathcal
      V}\gamma \in \mathcal V_h$ such that $d^\pm\sem{R}_{\mathcal
      V}\gamma = \sem{d^\pm\Xi}_{\mathcal V}\gamma$. We also extend
    this in the obvious way to transformation contexts
    $\sem{\Phi}_{\mathcal V}\gamma$
  \item For any $\Gamma\pipe\Phi \vdash t : R$ and $\gamma \in
    \sem{\Gamma}_{\mathcal V}$, a 2-cell $\sem{t}_{\mathcal V}\gamma
    \in V_h(\sem{\Phi}_{\mathcal V}\gamma;\sem{R}_{\mathcal V}\gamma)$
    with $d^\pm\sem{t}_{\mathcal V}\gamma = \id$.
  \end{itemize}

  %% \begin{align*}
  %%   \sem{\cdot}_{\mathcal V} &= 1\\
  %%   \sem{\Gamma, x:A}_{\mathcal V} &= \sum_{\gamma \in \sem{\Gamma}_{\mathcal V}} \sem{A}_{\mathcal V}\gamma\\
  %% \end{align*}

  %% with an 
  %% \begin{align*}
  %%   \sem{\Gamma, x:A}_{\mathcal V} &= \sem{\Gamma}_{\mathcal V} \times \sem{A}_{\mathcal V}\\
  %%   \sem{\Gamma}_{\mathcal V} &= 1\\\\
  %%   \sem{\Sort}_{\mathcal V} &= \mathcal V_0\\
  %%   \sem{\varr{\cat C}{\cat D}}_{\mathcal V} &= \mathcal V_v(\sem{\cat C})
  %% \end{align*}
\end{definition}

@@

>>>>

Next, we develop the basic model theory for SCT. Because SCT includes
not just a syntax for category theory, but is also indexed by an
ordinary context of dependent type theory $\Gamma$, a model of SCT is
not given by a \emph{single} virtual double category, but instead by
an \emph{indexed family} of VDCs, parameterized by an object of a
category that interprets dependent type theory. We call such a
structure as a \emph{VDC-hyperdoctrine}. We then show how to construct
a VDC-hyperdoctrine indexed by the category of sets, which has the
intuitive reading that any of the VDC judgments $\Gamma \vdash J$ is
interpreted as simply a family of $J$-data parameterized by elements
of the set $\Gamma$.

%% The first component of this is \emph{soundness}: every category,
%% object, set and transformation denote an object, vertical arrow,
%% horizontal arrow and 2-cell, respectively in any representable virtual
%% equipment, such that all equations in the theory are satisfied.
%% %
%% We combine this with a set theoretic semantics for types $A$ and
%% contexts $\Gamma$ to get a simple semantics of our logic over a
%% virtual equipment.
%% %
%% The second component of this is the \emph{completeness} theorem:
%% informally, the theory contains every category, object, set and
%% transformation that \emph{must} exist in any virtual equipment, and
%% further equations between syntactic forms hold if and only if they
%% hold in every model.
%% %
%% While sounding more ambitious, the completeness theorem is
%% typically simpler as it just amounts to showing that the syntax itself
%% forms a virtual equipment.

@@

First, we recount the definition of a virtual double category.
\begin{definition}[Virtual Double Category]
  A (vertically strict) virtual equipment $\mathcal V$ consists of
  \begin{enumerate}
    \item A set $\mathcal V_0$ of objects.
    \item For each $\cat C, \cat D \in V_0$, a set $V_v(\cat C, \cat
      D)$ of vertical arrows, with composition and unit operations
      satisfying the category axioms.
    \item A set $V_h$ of horizontal arrows, with $d^+,d^- : V_h \to
      V_0$.

    \item For any vertical arrows $f \in V_v(\cat C, \cat D)$ and $g
      \in V_v(\cat E, \cat F)$ and horizontal arrow $R \in V_h$ with
      $d^+R = \cat F$ and $d^-R = \cat D$, a \emph{restriction}
      $R(f,g)$ with $d^+R = \cat C$ and $d^-R = \cat E$.

      This operation is furthermore associative and unital $R(f \circ
      f',g \circ g') = R(f,g)(f',g')$ and $R(\id,\id) = R$.

    \item For each composable string $\Phi$ (defined below) of horizontal arrows
      starting at $\cat C$ and ending at $\cat D$, and horizontal
      arrow $R$ with $d^-R = \cat C$ and $d^+R = \cat D$, a set
      $V_2(\Phi; R)$ of 2-cells.

    \item For any string of vertical arrows $\vec f$ into $\Phi$ and
      2-cell $r \in \mathcal V_2(\Phi,S)$, a restriction operation
      $r(\vec f) \in \mathcal V_2(\Phi(\vec f), S(d^-(\vec f),
      d^+(\vec f)))$. Furthermore, this is associative and unital
      $r(\vec id) = r$ and $r(\vec f \circ \vec g) = r(\vec {f \circ
        g})$

    \item For each $R \in V_h$, an identity arrow $\id_R \in
      V_2(d^-R,R,d^+ R; R)$ with $d^\pm \id_R = \id_{d^\pm R}$.

    \item For any $2$-cell $\alpha \in V_2(\Phi;R)$ and composable
      string $\phi$ of 2-cells into $\Phi$, a composite $\alpha \circ
      \phi$.

    \item Composition should be unital in that:
      \[ \id_R \circ [d^-\alpha, \alpha, d^+\alpha] = \alpha  \]
      and
      \[ \alpha \circ [\vec \id] = \alpha\]
    \item And associative in that
      \[ (\alpha \circ [\beta_1,\ldots]) \circ [\gamma_{11},\ldots,\gamma_{21},\ldots,] = \alpha \circ [(\beta_1 \circ \gamma_{11},\ldots), (\beta_2 \circ \gamma_{21},\ldots)]\]
  \end{enumerate}

  A composable string $\Phi$ of horizontal arrows is defined to be an
  alternating sequence of objects and horizontal arrows, beginning and
  ending with an object such that any substring $\cat C,R,\cat D$
  satisfies $d^-R = \cat C$ and $d^+R = \cat D$.

  A composable string of vertical arrows $\vec f$ into $\Phi$ is a
  sequence of vertical arrows, one for each object in $\Phi$, with the
  codomain of the arrow matching the object in $\Phi$. Then

  A composable string of 2-cells from $\Phi$ to $\Psi$ is a sequence
  $\vec r$ of 2-cells, one for each horizontal arrow in $\Psi$ such
  that the codomains match the objects in
\end{definition}

Our main departure from\cite{crutwellshulman} is making restrictions
part of the definition, and also requiring them to be strictly
associative and unital. In the examples that we are most concerned
with, however, these operations are strict and so we will not explore
the possibility of coherence theorems for proving this equivalent to a
weaker definition.

\begin{definition}
  A functor of virtual double categories $F : \mathcal V \to \mathcal
  U$ is a mapping on objects, vertical arrows, horizontal arrows and
  2-cells respecting typing and preserving composition, units, and
  restriction.
\end{definition}

@@

\begin{definition}{Indexed Virtual Equipment}
  A VE-hyperdoctrine is a pair of a category $C$ and a functor $\cat
  C^{op} \to VE$. A morphism of VE-hyperdoctrines
  \[(F,\phi) : (\cat C, \mathcal V : \cat C^{o} \to VE) \to (\cat D, \mathcal U : \cat D^{o} \to VE)\]
  is a pair of a functor $F : \cat C \to \cat D$ and a natural
  transformation $\phi : \mathcal V \Rightarrow\mathcal U \circ F^{o}$.
\end{definition}

\begin{definition}{Soundness}

\begin{definition}{Completeness}
  We define the syntactic VE-hyperdoctrine $\textrm{Syn}$ as follows:
  \begin{enumerate}
  \item The base category is the category of contexts $\Gamma$ and substitutions.
  \item Define $\mathcal V_\Gamma$ as follows
    \begin{enumerate}
    \item An object of $\mathcal V^\Gamma$ is a proof $\Gamma \vdash \cat C \isaCat$.
    \item A vertical morphism $\mathcal V^\Gamma(\cat C, \cat D)$ is a
      term $\Gamma\pipe \alpha:\cat C \vdash t : \cat D$ moduleo
      $\alpha$-equivalence. Composition and identity are given by composition/variable.
    \item A horizontal morphism $R \in \mathcal^\Gamma_h$ is a term
      $\Gamma\pipe \alpha:\cat C;\beta:\cat D \vdash R$ with $d^-R =
      \cat C$ and $d^+R = \cat D$.
    \item Restriction of $R$ along $s,t$ is given by substitution
      $R[s/\alpha;t/\beta]$. This is clearly associative and unital.
    \item A 2-cell from $\Phi$ to $A$ is a term $\Phi \vdash t : B$
    \end{enumerate}
  \end{enumerate}
\end{definition}

%% \begin{definition}
%%   Given a well-formed signature $\Gamma$ and a virtual equipment
%%   $\mathcal V$, we mutually define the following notions:

%%   \begin{itemize}
%%   \item The set $\sem{\Gamma}_{\mathcal V}$
%%   \item For any $\Gamma \vdash A$ and $\gamma \in \sem{\Gamma}_{\mathcal V}$, a set $\sem{A}_{\mathcal V} \gamma$
%%   \item For any $\Gamma \vdash M : A$ and $\gamma \in \sem{\Gamma}_{\mathcal V}$, an element $\sem{M}_{\mathcal V} \in \sem{A}_{\mathcal V} \gamma$
%%   \item For any $\Gamma \vdash \cat C \isaCat$ and $\gamma \in \sem{\Gamma}_{\mathcal V}$, an object $\sem{\cat C}_{\mathcal V}\gamma$
%%   \item For any $\Gamma\pipe\alpha:\cat C \vdash a :\cat D$ and $\gamma \in \sem{\Gamma}_{\mathcal V}$, a vertical morphism $\sem{\alpha.a}_{\mathcal V}\gamma \in \mathcal V_v(\sem{\cat C}_{\mathcal V}\gamma, \sem{\cat D}_{\mathcal V}\gamma,)$
%%   \item For any $\Gamma\pipe\Xi \vdash R \isaSet$ and $\gamma \in
%%     \sem{\Gamma}_{\mathcal V}$, a horizontal arrow $\sem{R}_{\mathcal
%%       V}\gamma \in \mathcal V_h$ such that $d^\pm\sem{R}_{\mathcal
%%       V}\gamma = \sem{d^\pm\Xi}_{\mathcal V}\gamma$. We also extend
%%     this in the obvious way to transformation contexts
%%     $\sem{\Phi}_{\mathcal V}\gamma$
%%   \item For any $\Gamma\pipe\Phi \vdash t : R$ and $\gamma \in
%%     \sem{\Gamma}_{\mathcal V}$, a 2-cell $\sem{t}_{\mathcal V}\gamma
%%     \in V_h(\sem{\Phi}_{\mathcal V}\gamma;\sem{R}_{\mathcal V}\gamma)$
%%     with $d^\pm\sem{t}_{\mathcal V}\gamma = \id$.
%%   \end{itemize}

%%   %% \begin{align*}
%%   %%   \sem{\cdot}_{\mathcal V} &= 1\\
%%   %%   \sem{\Gamma, x:A}_{\mathcal V} &= \sum_{\gamma \in \sem{\Gamma}_{\mathcal V}} \sem{A}_{\mathcal V}\gamma\\
%%   %% \end{align*}

%%   %% with an 
%%   %% \begin{align*}
%%   %%   \sem{\Gamma, x:A}_{\mathcal V} &= \sem{\Gamma}_{\mathcal V} \times \sem{A}_{\mathcal V}\\
%%   %%   \sem{\Gamma}_{\mathcal V} &= 1\\\\
%%   %%   \sem{\Sort}_{\mathcal V} &= \mathcal V_0\\
%%   %%   \sem{\varr{\cat C}{\cat D}}_{\mathcal V} &= \mathcal V_v(\sem{\cat C})
%%   %% \end{align*}
%% \end{definition}

@@

\section{Example Models}

Our main intended models are when the indexing category is $\Set$ and
the hyperdoctrine is given by families of objects, arrows etc, of a
fixed virtual equipment $\mathcal V$, indexed by the set
$\sem\Gamma$. We can formalize this as the ``internal logic'' of the
virtual equipment $\mathcal V$.

\begin{definition}[Powers of Virtual Equipments]
  Let $\mathcal V$ be a virtual equipment and let $X$ be a set.
  Define a virtual equipment, the ``power'' $\mathcal V^X$ as follows.

  \begin{enumerate}
  \item Objects are functions $X \to \mathcal V_0$.
  \item Vertical morphisms $\mathcal V^X_v(A,B)$ are functions
    $\Pi_{x\inX} V_v(A x,B x)$. With pointwise composition, identity.
  \item Horizontal morphisms $\mathcal V^X_h$ are functions $\Gamma \to \mathcal V_h$, with
    $(d^\pm R)x = d^\pm (Rx)$.
  \item Restriction of $R$ along $f,g$ is given by $(R(f,g))x = (R x)(f x,g x)$.
  \item A 2-cell from $\Phi$ to $R$ is a family $\Pi_{x\in X} \mathcal
    V^X_h(\Phix,R x)$ with composition given pointwise.
  \end{enumerate}
\end{definition}

  \begin{enumerate}
  \item The indexing cateogry is $\Set$.
  \item The equipment is defined as $\mathcal U\Gamma = \mathcam V^\Gamma$.
  \item Restriction by a function $\gamma : \Delta \to \Gamma$ is
    given pointwise.
  \end{enumerate}
\end{definition}
